---
title: "wineQuality"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE,
                      fig.width=9, fig.height=5, fig.path='Figs/',
                      fig.align='center', tidy=TRUE)
```

#Data Analyst Nanodegree, Term 2 Project 2: EDA

Jesse Fredrickson

10/23/18

The purpose of this project is to perform exploratory data analysis (EDA) on a
dataset - I have elected to analyze a dataset involving the chemical properties
of white wine. I will do some basic analysis on select variables, and try to
investigate the relationships between wine properties and wine quality as judged 
by professional wine tasters. I will start by loading the dataset and doing some
basic 1 variable plots and analysis to understand the data.

```{r echo=FALSE}
# Load data
# setwd('C:/Users/1125193/Documents/PyData/DAND_t2_p2')
setwd('C:/Users/Jesse/Documents/PyData/DAND_t2_p2')
wq = read.csv('wineQualityWhites.csv')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# I will use the following packages in my EDA
# install.packages("ggplot2")
# install.packages("GGally")
# install.packages('tidyverse')
# install.packages('dplyr')
# install.packages('RColorBrewer')
# install.packages('gridExtra')
# install.packages('grid')
# install.packages('memisc')

library(ggplot2)
library(GGally)
library(RColorBrewer)
library(gridExtra)
library(grid)
library(memisc)
library(dplyr)
```

```{r echo=FALSE}
# First, get some structural information on the dataset
str(wq)
```

#Initial Observations
1) The dataset consists of almost 5000 observations, and 13 variables
2) The X variable appears to be an index number, I will ignore it most cases
3) I will likely want to compare everything to quality
4) Need to do research on acids, sugars, chlorides, and sulfurs + sulfites in
winemaking to get some concept of how they relate
5) May need to investigate collinearity between some variables
6) Although there are 3 variables pertaining to acidity, one is an order of
magnitude higher than the others and it may not make sense to combine them in
any sense
7) The quality variable is integers (continuous), and I will want to factor it
for some plots in order to use it as a categorical variable

#Winemaking Research
Sulfur Dioxide [mg/L]: is added to preserve wine due to its antimicrobial
properties. It is also a product of fermentation. Total sulfer dioxide is free +
bound, where bound sulfur dioxide has exhausted its preservative usefulness and
bound to sugars, acetaldehyde, and phenolic compounds.

Acidity [g/L]: Acids change the pH of the wine, cut the taste of sugar, and
enhance the effectiveness of sulfur dioxides. Fixed acidity refers to Tantric
acid, Volatile acidity mainly refers to Acetic acid, and Citric acidity of
course refers to citric acid. It is desired to have a balance of acidity and
sweetness (residual sugar).

Residual Sugar [g/L]: lends sweetness to wine (cuts dryness). Sugar is a reagent
in the fermentation process by which yeast produces alcohol.

Chlorates [g/L]: refers to Sodium Chloride, analogous to salt in wines.

Sulphates [g/L]: refers to Potassium Sulphate.

Density: Alcohol is less dense than sugar, therefore as sugar is consumed during
fermentation and alcohol is produced, the density goes down.

#Univariate Plots

The variable I care most about is quality. I will first look at the distribution
of quality rankings

```{r echo=FALSE}
# Plot histogram of qualities
ggplot(data = wq, aes(x = quality)) +
  geom_histogram(binwidth = 1)
```

```{r echo=FALSE}
summary(wq$quality)
```

Wine quality appears to follow a normal distribution, centered around rating of
6. There are only 7 wine ratings that are actually in use between 3 and 9. This
distribution is good because I should have enough data points to differentiate 
between poor quality (<6) and good quality (>6) wines.

Next I will create a histogram of alcohol content.

```{r echo=FALSE}
# Histogram of alcohol content in wines
ggplot(data = wq, aes(x = alcohol)) +
  geom_histogram(binwidth = .2)
```

The distribution of alcohol is left skewed, meaning the majority of wines have
an alcohol content between 9-11%, but has a long tail with some wines having an
alcohol content of 14%. There does appear to be some bimodality going on with
one peak at around 9.5% and a second peak at 12.25%.

I will show one more histogram on density, since it is another easy to
understand variable.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Histogram of wine density
wq$quality.ordered <- ordered(wq$quality)

ggplot(data = wq, aes(x = density)) +
  geom_histogram(bins=70) +
  xlim(quantile(wq$density, .001), quantile(wq$density, .999))
```

Again we see a nice distribution, appearing fairly normal. There does appear to
be a pair of peaks on the right half of the distribution, but it is not clear if
they are significant.

Next I will try examining residual sugar.

```{r echo=FALSE}
ggplot(data = wq, aes(x = residual.sugar)) +
  geom_histogram(bins = 100) +
  xlim(0, quantile(wq$residual.sugar, probs = .99))
```

Here we see a very long tailed left skewed distribution. It appears that many 
wines have very low sugar.

I will also be interested in the acidity of wines.

```{r echo = FALSE, message=FALSE, warning=FALSE}

p1 <- ggplot(data = wq, aes(x = citric.acid)) +
  geom_histogram(bins = 100) +
  xlim(0, quantile(wq$citric.acid, probs = .99))

p2 <- ggplot(data = wq, aes(x = volatile.acidity)) +
  geom_histogram(bins = 100) +
  xlim(0, quantile(wq$volatile.acidity, probs = .99))

p3 <- ggplot(data = wq, aes(x = fixed.acidity)) +
  geom_histogram(bins = 100) +
  xlim(quantile(wq$fixed.acidity, probs = .01),
       quantile(wq$fixed.acidity, probs = .99))

grid.arrange(p1,p2,p3)
```

Each of these plots follows a normal distribution, excepting citric acid, which
has a large spike at approximately .5.

Similarly with pH

```{r}
ggplot(data = wq, aes(x = pH)) +
  geom_histogram()
```

Again, a nice normal distribution.

Finally, I will be interested in free and total sulfur dioxide.

```{r}
p1 <- ggplot(data = wq, aes(x = free.sulfur.dioxide)) +
  geom_histogram(bins = 50) +
  xlim(0, quantile(wq$free.sulfur.dioxide, probs = .99))

p2 <- ggplot(data = wq, aes(x = total.sulfur.dioxide)) +
  geom_histogram(bins = 50) +
  xlim(0, quantile(wq$total.sulfur.dioxide, probs = .99))


grid.arrange(p1,p2)
```

Each has a normal distribution. Free Sulfur Dioxide has curious breaks in its
bins, but this may be an artifact of the binning. If I reduce the number of bins
the breaks disappear, but I kept it as is in case this  behavior becomes
important in the future.

#Univariate Analysis
In this section I found that the data involves 11 continuous variables and 1
factored variable (quality). I will mainly be interested in the quality
variable and what influences it. I added 1 column to the dataset,
quality.ordered, which is a faceted version the quality variable, which was just
an integer in its original state. I will also sum the different acids to see if
they have a cumulative effect which is not otherwise obvious. Of the variables
which I plotted above, all fit fairly normal curves, which is encouraging
because an unusual distribution could signify the effect of an external variable
which might not be accounted for by chemistry alone.

#Bivariate Plots
I will start with a scatter matrix, comparing all variables against one another
and calculating correlation coefficients in one step.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# scatter matrix, less x
ggpairs(data = dplyr::select(wq[sample(nrow(wq), 1000), ], -X, -quality.ordered),
        progress=FALSE) +
  theme(text = element_text(size=7)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 7))
# ggsave('wineQuality_scatter.png')
```

It is necessary to zoom in on this plot in a new window in order to use it. See 
the attached wineQuality_scatter.png for details. The scatterplot matrix is
useful because it immediately reveals relationships between all variable pairs
in a dataframe, while also yielding line plot histograms of each solo variable.
For instance, we can see that fixed acidity has a nice normal curve, white
citric acidity is bimodal. Chlorides appears normal with a few high outliers,
and resigual sugars is left skewed. The plots I am most interested in are on the
last line, which are each variable versusquality. No individual variable has a
strong correlation with quality, but density and alcohol do appear to have some
weak correlation.

I will first look at density versus quality with two plots.

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = wq, aes(x = density)) +
  geom_histogram(aes(fill = quality.ordered)) +
  xlim(quantile(wq$density, .001), quantile(wq$density, .999))
```

```{r echo=FALSE}
# Wine density vs quality, revisualized
# jitter density vs quality: weak relationship?
ggplot(data = subset(wq, density < quantile(density, probs = .99)),
       aes(x = density, y = quality)) +
  geom_point(alpha = .20, position = position_jitter())
```

Here we see the relationship between density in two plots. As I hypothesized,
there appears to be a weak relationship between density and quality. From my
research I know that a low density is associated with a high alcohol content, so
I will look to see if the same behavior exists with alcohol content.

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = wq, aes(x = alcohol)) +
  geom_histogram(aes(fill = quality.ordered)) +
  xlim(quantile(wq$alcohol, .001), quantile(wq$alcohol, .999))
```

```{r echo=FALSE}
# plot alcohol vs quality: weak relationship
ggplot(data = wq,
       aes(x = alcohol, y = quality)) +
  geom_point(alpha = .20, position = position_jitter())
```

As expected, higher alcohol content does appear to be associated with higher
quality ratings. The first chart may be more revealing than the second in that
we see a few higher quality ratings even though the number of high alcohol
content samples is lower! Another way of visualizing this relationship is with a
boxplot, as follows.

```{r echo=FALSE}
ggplot(data = wq,
       aes(x=quality.ordered, y=alcohol)) +
  geom_boxplot()
```

This plot shows more clearly the relationship between alcohol and quality, 
although it misses the information about the distribution of quality. Starting
the worst quality wine, as we move up in quality the median alcohol content
drops by about 1% ABV until the wine reaches a quality of 6, where the ABV rises
to 12.5% ABV at a quality of 9.

One more relationship that was revealed by the scatter matrix was a strong
correlation (0.84) between residual sugar and density.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wq,
       aes(x = residual.sugar,
           y = density)) +
  geom_point(alpha = .25) +
  geom_smooth() +
  xlim(0, quantile(wq$residual.sugar, probs = .99)) +
  ylim(quantile(wq$density, probs = .01), quantile(wq$density, probs = .99))
```

It should not be surprising that as residual sugar rises, so does density, as
sugar is what is consumed in the fermentation process to produce alcohol, and
alcohol has a low density.
  
Next I will see if I can flesh out more complex relationships, particularly
relating to quality

#Bivariate Analysis
In this section I found that no supporting variables have a particularly strong
relationship with the main variable, quality. However, both density and alcohol
content do correlate weakly with quality. Low density and high alcohol content
are both associated with higher quality ratings on average. The scatter matrix
also revealed a strong correlation between density and residual sugar, which I
replotted with a smooth curve to demonstrate the relationship.

#Multivariate Plots
The scatter matrix showed that none of the variables for acidity had a strong
relationship directly with quality, but I might be able to derive on if I can 
find a relationship between those supporting variables themselves. 
I know that pH is a measure of acidity, so I would expect the pH variable to
correlate somehow with the amount of acid in a wine. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
#plot each type of acidity
grid_arrange_shared_legend <- function(...) {
    plots <- list(...)
    g <- ggplotGrob(plots[[1]] + theme(legend.position="bottom"))$grobs
    legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
    lheight <- sum(legend$height)
    grid.arrange(
        do.call(arrangeGrob, lapply(plots, function(x)
            x + theme(legend.position="none"))),
        legend,
        ncol = 1,
        heights = unit.c(unit(1, "npc") - lheight, lheight))
}


p1 <- ggplot(data = wq,
            aes(x = fixed.acidity,
                y = pH,
                color = quality.ordered)) +
  geom_point(position = position_jitter()) +
  xlim(quantile(wq$fixed.acidity, probs = .01),
       quantile(wq$fixed.acidity, probs = .95)) +
  scale_color_brewer(type = 'div')+
  theme_dark()

p2 <- ggplot(data = wq,
            aes(x = volatile.acidity,
                y = pH,
                color = quality.ordered)) +
  geom_point(position = position_jitter()) +
  xlim(quantile(wq$volatile.acidity, probs = .01),
       quantile(wq$volatile.acidity, probs = .95)) +
  scale_color_brewer(type = 'div')+
  theme_dark()

p3 <- ggplot(data = wq,
            aes(x = citric.acid,
                y = pH,
                color = quality.ordered)) +
  geom_point(position = position_jitter()) +
  xlim(quantile(wq$citric.acid, probs = .01),
       quantile(wq$citric.acid, probs = .95)) +
  scale_color_brewer(type = 'div')+
  theme_dark()

grid_arrange_shared_legend(p1,p2,p3)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# plot cum acidity vs pH vs quality: no relationship

# make total acidity column
wq$total.acidity <- with(wq, fixed.acidity + volatile.acidity + citric.acid)

ggplot(data = wq,
       aes(x = total.acidity, y=pH, color = quality.ordered)) + 
  geom_point(alpha = 1, position = position_jitter()) +
  xlim(quantile(wq$total.acidity, probs = .01),
       quantile(wq$total.acidity, probs = .95)) +
  ylim(2.5,4) +
  scale_color_brewer(type = 'div') +
  theme_dark()
```

As expected, we see that as total acidity increases, pH decreases slightly.
There does not appear to be any correlation between pH and quality, or acidity
and quality. Total Acidity is affected primarily by fixed acidity, and very
little by volatile or citric acidity.

I would also expect there to be a relationship between free and total sulfur
dioxide. I will repeat the above colored scatter plot for total.sulfur.dioxide
and free.sulfur.dioxide versus quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# experimenting with byvariate scatter plots colored by quality
# There is a relationship between free and total sulfur dioxide
ggplot(data = wq,
       aes(x = total.sulfur.dioxide,
           y=free.sulfur.dioxide,
           color = quality.ordered)) + 
  geom_point(alpha = 1, position = position_jitter()) +
  scale_color_brewer(type = 'div') +
  xlim(0,quantile(wq$total.sulfur.dioxide, probs = .95)) +
  ylim(0,quantile(wq$free.sulfur.dioxide, probs = .95))+
  theme_dark()
```

There appears to be a 'sweet spot' for free sulfur dioxide and total sulfur
dioxide where one area between 30-50 mg/L free sulfur dioxide and 75-125 mg/L
total sulfur dioxide has a higher density of 7+ quality ratings. I will indlude
these variables in a model I make later to see if they lend themselves to
predicting quality because of this.

The strongest correlation between any two variables on the scatter matrix was
between density and residual sugar, which makes sense because as sugar is
consumed by the fermentation process it will be converted into alcohol, which
has a lower density. Knowing this, I will see if there is any correlation
between density and residual.sugar vs quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wq,
       aes(y = density, x=residual.sugar, color = quality.ordered)) + 
  geom_point(alpha = 1, position = position_jitter()) +
  scale_color_brewer(type = 'div') +
  ylim(quantile(wq$density, probs = .05),
       quantile(wq$density, probs = .95)) +
  xlim(quantile(wq$residual.sugar, probs = .05),
       quantile(wq$residual.sugar, probs = .95))+
  theme_dark()
```

Here we finally see a complex relationship. As expected, as residual sugar
rises, so does density. But what we also see is that if we hold residual sugar
constant and raise the density, quality goes down. This could imply that there
is a imbalance in sugar and alcohol, resulting in a wine that is too sweet.
Let's try repeating the plot with alcohol instead of density to see what that
yields. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wq,
       aes(y = alcohol, x=residual.sugar, color = quality.ordered)) + 
  geom_point(alpha = 1, position = position_jitter()) +
  scale_color_brewer(type = 'div') +
  ylim(quantile(wq$alcohol, probs = .01), quantile(wq$alcohol, probs = .99)) +
  xlim(quantile(wq$residual.sugar, probs = .01), quantile(wq$residual.sugar, probs = .99))+
  theme_dark()
```

This plot shows what we saw before, that higher alcohol contents are often rated
higher quality. But it does not reveal any deeper trends other than that very
few wines have both high residual sugar and high alcohol content.

Using some of the variables I have used already, I will try generating a model
and seeing how well it can explain variations in quality.

```{r echo=FALSE}
m1 <- lm(quality ~ alcohol, data = wq)
m2 <- update(m1, ~ . + density)
m3 <- update(m2, ~ . + residual.sugar)
m4 <- update(m3, ~ . + free.sulfur.dioxide)
m5 <- update(m4, ~ . + total.sulfur.dioxide)
m6 <- update(m5, ~ . + pH)
mtable(m1, m2, m3, m4, m5, m6)
```

Evidently, the linear model does a very poor job of fitting the dataset. The
correlation coefficient using 6 variables was only .229. Interestingly, the best
predictor remained to be alcohol, which has a .190 correlation coefficient by
itself.

#Multivariate Analysis
The primary relationships observed in the multivariate cases were those between
density, alcohol, residual sugar, and quality. As expected, alcohol is inversely
correlated with density, which is positively correlated with residual sugar. The
relationship I was not expecting was between free and total sulphur dioxide and
quality, where there appeared to be a 'sweet spot' for quality. Again this could
have to do with a 'balance' of some kind, or it could be related to some other
variable which I did not investigate.

The 6 degree linear model I created did a very poor job fitting the quality
variable, which suggests that even among variables that exhibited some kind of 
a relationship with quality, the relationship is nonlinear or incomplete.

#Final Plots and Summary

### First Plot

```{r echo=FALSE}
give.n <- function(x){
   return(c(y = median(x) +0.125, label =length(x)))
}

ggplot(data = wq,
       aes(x=quality.ordered, y=alcohol)) +
  geom_boxplot() +
  ylab('Alcohol [%]') +
  xlab('Quality Rating') +
  ggtitle('Alcohol Content by Quality Rating (count shown)') +
  stat_summary(fun.data = give.n,
               geom = "text",
               position = position_dodge(width = 1),
               size = 3)
```

### Description 1
This plot shows the median and IQR of alcohol content per quality rating, and I
have also added the count of samples per rating to each box. It is evident that
the highest rated wines have the highest alcohol content, but that the opposite
end of the scale also shows a lift in alcohol ratings, possibly because the
lowest end wines are intended to be strong if not tasty. Meanwhile the high end
options are ones that can blend taste with strength. Or perhaps the raters just
rate higher when they are inebriated.

### Second Plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wq,
       aes(y = density, x=residual.sugar, color = quality.ordered)) + 
  geom_point(alpha = 1, position = position_jitter()) +
  scale_color_brewer(type = 'div', name = 'Quality') +
  ylim(quantile(wq$density, probs = .05),
       quantile(wq$density, probs = .95)) +
  xlim(quantile(wq$residual.sugar, probs = .05),
       quantile(wq$residual.sugar, probs = .95)) +
  xlab('Residual Sugar [g/L]') +
  ylab('Density [kg/L]') +
  ggtitle('Residual Sugar vs Density vs Quality')+
  theme_dark()

```

### Description 2
Here we see a relationship that is only clear on a 3 variable plot like this
one. Density and residual sugar are both correlated with quality, but only when
the other is held constant, because density and residual sugar are also
positively correlated with one another. My interpretation of this plot is that
for a given amount of residual sugar, removing alcohol (and raising density)
will yield a wine that is too sweet, lowering its quality rating.

### Third Plot

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wq,
       aes(x = total.sulfur.dioxide,
           y=free.sulfur.dioxide,
           color = quality.ordered)) + 
  geom_point(alpha = 1, position = position_jitter()) +
  scale_color_brewer(type = 'div', name = 'Quality') +
  xlim(0,quantile(wq$total.sulfur.dioxide, probs = .95)) +
  ylim(0,quantile(wq$free.sulfur.dioxide, probs = .95)) +
  xlab('Total Sulfur Dioxide [mg/L]') +
  ylab('Free Sulfur Dioxide [mg/L]') +
  ggtitle('Sulfur Dioxide Free, Total, vs Quality')
```

### Description 3
This plot shows the relationship of total and free sulfur dioxide. As expected,
total sulfur dioxide rises with free sulfur dioxide. What is interesting about
this plot is that there is a 'sweet spot' in terms of wine quality right
in the middle of the plot. It is unclear to me why this is, and I was not able
to uncover strong relationships between sulfur and other variables in this
dataset, so that is something that I would like to investigate further in the
future.

#Reflection
This was a challenging dataset, because as was made clear by the scatter matrix,
there are very few ostensible correlations between individual variables. Through
some pseudo-random experimentation I was able to flesh out some interesting
interactions, particularly relating to density, alcohol, residual sugar, total
and free sulfur dioxide, and quality, but I was not able to tie them all
together in an effective model. What was surprising to me was that of all
variables, it seems that alcohol content by itself was easily the best predictor
of quality. It would have been useful to have more info on wines beyond chemical
composition, such as year, vineyard, price etc. Without more info to go on, my
best advice when wine shopping is just to buy the strong stuff. According to
this dataset, if you aim for >= 12.8% ABV, you are more likely than not to get 
above average quality wine!


